---
title: "Introduction to R"
format: html
---

## What is R?

R is a free, open-source programming language and environment specifically designed for statistical computing and data analysis. What makes R truly powerful is its **modular design** - R consists of a core system that provides basic functionality, which is then extended through thousands of specialized packages contributed by statisticians, data scientists, and researchers worldwide.

## The package ecosystem: R's superpower

R's strength lies in its package system. Think of R as a smartphone: the base R installation is like the phone's operating system, providing essential functions. Packages are like apps - each one adds specific capabilities for different tasks.

### Base R vs. packages

**Base R** includes fundamental functions for:
- Basic arithmetic and statistics
- Data structures (vectors, data frames, lists)
- Simple graphics
- File input/output

**Packages** extend R with specialized tools for:
- Advanced statistical methods
- Machine learning algorithms
- Data visualization
- Web scraping
- Bioinformatics
- Finance and economics
- PX-files
- And much more!

### CRAN: The package repository

The Comprehensive R Archive Network (CRAN) hosts over **19,000 packages**. This massive ecosystem means that whatever data analysis task you're facing, someone has likely created a package to help.

```{r}
#| eval: false
# Install packages from CRAN
install.packages("tidyverse")    # Collection of data science packages
```

## Key package collections

### The tidyverse

The [tidyverse](https://www.tidyverse.org/) is a collection of packages designed for data science with a consistent philosophy and grammar:

```{r}
#| eval: false
library(tidyverse)

# This loads 8 core packages:
# - ggplot2: data visualization
# - dplyr: data manipulation
# - tidyr: data tidying
# - readr: data import
# - purrr: functional programming
# - tibble: modern data frames
# - stringr: string manipulation
# - forcats: working with factors
```

R has package ecosystems for virtually every field. So if you need a package in a specific field, browse the web to find it.


## Basic R concepts (built on packages)

### Data structures

Even basic data operations benefit from packages:

```{r}
# Base R data frame
students_base <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 22, 19),
  grade = c(85, 92, 78)
)

# With tibble (tidyverse package) - enhanced data frames
library(tibble)
students_tibble <- tibble(
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 22, 19),
  grade = c(85, 92, 78)
)

students_tibble  # Better printing and behavior
```

### Data manipulation

Base R can manipulate data, but packages make it easier:

```{r}
#| eval: false
# Base R
high_grades <- students_base[students_base$grade > 80, ]
mean_age <- mean(students_base$age)

# With dplyr (tidyverse package) 
library(dplyr)
high_grades <- students_tibble %>% 
  filter(grade > 80)

mean_age <- students_tibble %>% 
  summarise(avg_age = mean(age))
```

## Package management

### Installing and loading packages

```{r}
#| eval: false
# Install once per computer
install.packages("dplyr")

# Load each R session
library(dplyr)

# Alternative loading method
require(dplyr)

# Install multiple packages at once
install.packages(c("dplyr", "ggplot2", "readr"))
```


## Why the package approach works

1. **Specialization**: Experts in each field contribute domain-specific tools
2. **Quality control**: CRAN has submission standards ensuring package quality
3. **Community-driven**: Thousands of contributors improve and maintain packages
4. **Modularity**: Only load what you need, keeping R fast and efficient
5. **Innovation**: New methods are quickly available as packages

## Getting started workflow

1. **Install base R** and RStudio
2. **Identify your needs** (data visualization, statistical modeling, etc.)
3. **Find relevant packages** using CRAN Task Views or online resources
4. **Install and explore** packages with documentation and examples
5. **Combine packages** to build powerful analysis workflows

R's package-centric design means you're never starting from scratch - you're building on the work of thousands of experts who've solved similar problems. This collaborative approach makes R incredibly powerful for data analysis across virtually any domain!