---
title: "Validating PX-files via API"
format: html
editor_options: 
  chunk_output_type: console
---

## The `pxweb` package

The `pxweb` package has functions to work with PX-web APIs. It works with any PX-web database that has the API feature enabled. Below we collect a table from the [Ghanaian StatsBank](https://statsbank.statsghana.gov.gh/pxweb/en/PHC%202021%20StatsBank/).

```{r}
#| message: false

library(pxweb)
library(tidyverse)

from_api <- pxweb_get(url = "https://statsbank.statsghana.gov.gh:443/api/v1/en/PHC 2021 StatsBank/Education and Literacy/attended_table.px",
                      query = list(
                        Highest_level_of_education = c("*"),
                        Geographic_Area = c("Ghana"),
                        Locality = c("*"),
                        Sex = c("*"),
                        Age = c("All ages", "15-17 years", "18 years & older"))) %>% 
  as.data.frame() %>% 
  as_tibble() # as_tibble for nicer presentation

head(from_api) %>% 
  kableExtra::kable()

```

If we want to update this table with data for a new year, we can use data from the API to validate that our data for the new year looks as expected. Now we create some new random data from the dataset for a new year.

```{r}
#| code-fold: true
#| code-summary: generate random data

set.seed(42); new_year <- from_api %>% 
  mutate(`Population (15 years and older)_new_year` = sample(`Population (15 years and older)`),
         .keep = "unused")

```

Now we can join the data together and check for developments.

```{r}

pct_change <- from_api %>% 
  left_join(new_year,
            by = join_by(Highest_level_of_education, 
                         Geographic_Area, Locality, 
                         Sex, Age)) %>% 
  mutate(pct_change = (`Population (15 years and older)_new_year` - `Population (15 years and older)`)/
           `Population (15 years and older)`*100)

```

Now we can use this new column `pct_change` to see if we have some unexpectedly high changes (above 10%), which might indicate problems in our code generating the data.

```{r}

if(max(abs(pct_change$pct_change), na.rm = TRUE) > 10){
    warning("There are changes above 10% for the new year. This may indicate problems with the data")
  print(pct_change %>% 
          filter(abs(pct_change) > 10)) %>% 
    kableExtra::kable()
} else {
  cat("No changes above 10%")
}

```

We see that we have some extremely high changes. For example, the value for the total across the variables has gone from 11,289,655 to 569. This clearly indicates a mistake. In this case, the mistake stems from our random data generation. We can instead do it in a more controlled process.

```{r}
#| code-fold: true
#| code-summary: Controlled data generation

set.seed(42); new_year <- from_api %>% 
  mutate(`Population (15 years and older)_new_year` = 
           `Population (15 years and older)` * runif(n(), 0.9, 1.1),
         .keep = "unused")

```

```{r}

pct_change <- from_api %>% 
  left_join(new_year,
            by = join_by(Highest_level_of_education, 
                         Geographic_Area, Locality, 
                         Sex, Age)) %>% 
  mutate(pct_change = (`Population (15 years and older)_new_year` - `Population (15 years and older)`)/
           `Population (15 years and older)`*100)

if(max(abs(pct_change$pct_change), na.rm = TRUE) > 10){
  print(pct_change %>% 
          filter(abs(pct_change) > 10))
  stop("There are changes above 10% for the new year. This may indicate problems with the data")
} else {
  cat("No changes above 10%")
}

```

Now we have no year-to-year changes above 10% (of course, because our data generation process set a limit of 10%). This limit of 10% is not set in stone and should be adjusted according to the content of the data. For example, some countries have seen inflation rates above 10%, so in that case the relevant number might be different.

Here we checked using data from the PX-web API, also to showcase the API. If the PX-file is available on disk, it may be easier to just load that and then do some validation checks.

The R package `validate` is specifically focused on tools and functions for validating data. For more information on data validation, see [The Data Validation Cookbook](https://cran.r-project.org/web/packages/validate/vignettes/cookbook.html), which introduces the R package `validate` and data validation concepts and implementations in R in general.

## Check cross-sums

Usually in a Statbank/PX-web database, we have multiple tables covering one topic, e.g., multiple tables to cover the theme of population statistics. Often these tables will share some variables (sex and age would be prime examples), and therefore we expect values to be the same across tables.

We have our data collected from the [Ghanaian StatsBank](https://statsbank.statsghana.gov.gh/pxweb/en/PHC%202021%20StatsBank/PHC%202021%20StatsBank__Education%20and%20Literacy/attended_table.px/) about education statistics for persons 15 years and older who attended school in the past.

We have another table about [Population (3 years and older) by School Attendance Status, District, Region, Type of Locality, Age and Sex](https://statsbank.statsghana.gov.gh/pxweb/en/PHC%202021%20StatsBank/PHC%202021%20StatsBank__Education%20and%20Literacy/sch_attend_stat_table.px/), and here we would expect that it shares some values with our table from earlier if we filter school attendance status to "attended in past" and age to over 15 years. Now we collect this data using the `pxweb` package.

```{r}

check_table <- pxweb_get(url = "https://statsbank.statsghana.gov.gh:443/api/v1/en/PHC 2021 StatsBank/Education and Literacy/sch_attend_stat_table.px",
                      query = list(
                        Schoolattendancestatus = c("Attended in the past"),
                        Geographic_Area = c("Ghana"),
                        Locality = c("*"),
                        Sex = c("*"),
                        Age = c("15-19", "20-24", "25-29", "30-34",
                                "35-39", "40-44", "45-49", "50-54",
                                "55-59", "60-64", "65-69", "70-74",
                                "75-79", "80-84", "85-89", "90-94",
                                "95-99", "100+"))) %>% 
  as.data.frame() %>% 
  mutate(Locality = if_else(Locality == "All locality types", 
                            "All Locality Types",
                            Locality)) %>% 
  as_tibble() # as_tibble for nicer presentation

head(check_table) %>% 
  kableExtra::kable()

```

This table we want to check uses another age definition, so we just need to sum up all the age groups to have persons over the age of 15.

```{r}
check_table_above_15 <- check_table %>% 
  summarise(`Population (3 years and older)` = sum(`Population (3 years and older)`),
            .by = c(Schoolattendancestatus, Geographic_Area, Locality, Sex))
```

Now we have the values for persons aged 15 and over, which would correspond to the value for all ages in our initial table. Let's select the corresponding variables between the tables and join them together.

```{r}

cross_check <- from_api %>%
  filter(Age == "All ages" & Highest_level_of_education == "Total") %>% 
  select(Geographic_Area, Locality, Sex, `Population (15 years and older)`) %>% 
  ## Joining data together
  left_join(check_table_above_15 %>% 
              select(Geographic_Area, Locality, Sex, `Population (3 years and older)`),
            by = join_by(Geographic_Area, Locality, Sex)) %>% 
  ## Calculating difference
  mutate(diff = `Population (15 years and older)` - `Population (3 years and older)`)
  
```


Now we have joined the two tables together and calculated a difference variable that hopefully should be 0.

```{r}
unique(cross_check$diff)
```

Which it is. The values shared across the two tables are the same, which is good as it indicates a consistent methodology in the two tables. This could also be a way to check a newly generated PX-table if you know that you have a PX-table in your database that shares some of the values.